---
title: "Test of XGBoost Regression: Predicting Physiology from Gene Expression"
author: "Anna Pardo"
date: "Jan. 29, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Brief Project Summary

Photosynthesis is central to plant carbon acquisition, but the typical pathway ("C3") is inhibited by abiotic stresses such as water limitation due to the dual function of stomata in carbon intake and transpiration. Crassulacean acid metabolism (CAM) is a form of photosynthesis in which stomata are opened mostly at night, when temperature is reduced, thus reducing transpiration and increasing the plant’s water use efficiency. CAM exists across a spectrum in many lineages, including in the genus *Yucca*; some species, such as *Y. filamentosa*, are strictly C3, whereas others, such as *Y. aloifolia*, use mainly the CAM pathway, and others use CAM only under certain conditions (“facultative CAM” or “C3+CAM”). One such species is *Y. gloriosa*, a homoploid hybrid of CAM *Y. aloifolia* and C3 *Y. filamentosa*, which exhibits intraspecific variation in the degree of facultative CAM under drought. The combination of homoploidy and unique CAM biology makes *Y. gloriosa* an excellent system for the study of both CAM evolution and homoploid speciation. To this end, a 24-hour timecourse RNA-seq dataset was collected for a controlled drought experiment run on both parent species and 26 naturally occurring genotypes of Y. gloriosa. These data are being leveraged to understand both the expression variability arising from homoploid hybridization and that underlying intraspecific CAM variation in this species.

### Purpose of This Analysis

This document contains code & results of a test XGBoost machine learning analysis to predict drought photosynthetic physiology data (carbon assimilation or `photo` and stomatal conductance or `cond`) from gene expression data in *Y. gloriosa*. The ultimate goal of this analysis is to use feature importance output from the model to find important genes for the facultative CAM phenotype in *Y. gloriosa*.

```{r load packages}
library(dplyr)
library(readr)
library(xgboost)
library(caret)
```

#### Load data

Both the X (TPM for all genes in *Y. gloriosa*) and the Y (`photo` or `cond` data paired with RNA-seq data) are contained in the same file:

```{r load data}
tpmphys <- as.data.frame(read_delim("//wsl.localhost/Ubuntu/home/leviathan22/yucca-genomics/machine_learning/paired_TPM_physiology.txt",delim = "\t"))
str(tpmphys)
```

#### Data subsetting

Since facultative CAM status is determined most by physiology under stress (in this case, drought `D`), data are filtered to keep only drought samples. Additionally, for this test model I will predict stomatal conductance, so I will drop the carbon assimilation column.

```{r filter to drought only}
dsamp <- tpmphys %>% filter(treat=="D") %>% select(-photo)
```

#### Log-transform TPM values

```{r logTPM}
tpmonly <- dsamp %>% select(-c(sample_name,genotype,treat,ZT,cond))
# for this non-HPC test: drop some columns/genes (try predicting on only the first 500 instead of the ~86,000 total that exist)
tpmonly <- tpmonly[,1:500]
# log-transform the subset data
logtpm <- tpmonly %>% mutate_all(~ log2(.+1))
# reattach the metadata columns
md <- dsamp %>% select(c(sample_name,genotype,treat,ZT,cond))
logtpm <- cbind(md,logtpm)
```

#### Metadata setup
Rather than doing random train/test split, I want to use certain genotypes as the training data and others as test data. This will avoid data leakage.

```{r metadata}
# plot: how many samples are present for each genotype?
sampcount <- md %>% count(genotype) %>% arrange(desc(n))

ggplot(sampcount,aes(x=genotype,y=n))+geom_col()+theme_bw()
```
Conclusion: We should remove genotypes 37, 43, and 48, which have only one sample each.

```{r drop genotypes}
logtpm <- logtpm %>% filter(!genotype %in% c("37","43","48"))
md <- md %>% filter(!genotype %in% c("37","43","48"))
```

What percentage of the remaining data does each genotype represent?

```{r genotype percentages}
sampcount <- sampcount %>% filter(!genotype %in% c("37","43","48"))
sampcount <- sampcount %>% ungroup() %>% mutate(percent = n / sum(n) * 100)

ggplot(sampcount,aes(x=genotype,y=percent))+geom_col()+theme_bw()
```

#### Split training & test data

I would like to preserve a roughly 80:20 train:test split, although my genotypes don't match that exactly. I will set up a function to pull random genotypes until the test data represent 15-25% of the total data.

```{r train test split function}
genotype_split_selection <- function(pctdf,minval=15,maxval=25){
  p=0
  gtselec <- c()
  while(p<minval) {
    gt <- sample(pctdf$genotype,size=1)
    pct <- pctdf %>% filter(genotype == gt) %>% pull(percent)
    p <- p + pct
    if(p<maxval){
      gtselec <- append(gtselec,gt)
      pctdf <- pctdf %>% filter(genotype != gt)
    } else {
      p <- p - pct
    }
  }
  print(paste("Test percentage:",as.character(p)))
  return(gtselec)
}
```

```{r train test split}
# select genotypes for test data
testgts <- genotype_split_selection(sampcount)

# set sample ID as index
rownames(logtpm) <- logtpm$sample_name

# split train & test data by genotype
trainData <- logtpm %>% filter(!genotype %in% testgts) %>% select(-c(sample_name,genotype,treat,ZT))
testData <- logtpm %>% filter(genotype %in% testgts) %>% select(-c(sample_name,genotype,treat,ZT))

# Separate the features and target variable
train_x <- as.matrix(trainData %>% select(-cond))
train_y <- trainData$cond
test_x <- as.matrix(testData %>% select(-cond))
test_y <- testData$cond
```

#### Hyperparameter tuning

```{r hyperparameter tuning}

```

#### Train the model & make predictions

Note, this is the naive version - will need to do hyperparameter tuning in the final version.

```{r xgboost}
# train the model
xgb_model <- xgboost(x = train_x, 
                     y = train_y, 
                     nrounds = 100, 
                     objective = "reg:squarederror")

# predict the values for the test set
preds <- predict(xgb_model, test_x)

# calculate RMSE
rmse <- sqrt(mean((preds - test_y)^2))
print(paste("RMSE:", rmse))
```
#### Plot actual vs. predicted values to evaluate the model
```{r plot values}
# combine preds & test_y into a dataframe
all_y <- as.data.frame(preds)
all_y$actual <- test_y
ggplot(all_y,aes(x=preds,y=actual))+geom_point() + theme_bw()

```

This plot shows that the predictions are not very good. What is the R-squared value?